Moisture absorption and drying alter nonadditive litter flammability in a mixed conifer forest

### 2020-04-02 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Combine figures as requested by reviews on R1

%% Revision 2019-12-10
\documentclass[letterpaper,12pt]{article}
\usepackage{jecol}

@@

\nwords{8200}
\ntables{4}
\nfig{8}
\nref{55}

@@

Fuel moisture content through time is controlled by two processes, wetting
(including absorption and adsorption) and drying rate (desorbtion)
\citep{Anderson-1990, Kreye+Kobziar+etal-2013}. Absorptive capacity represents
the potential maximum fuel moisture and is evaluated by the moisture content of
litter when saturated; hereafter we refer to this measurement as ``maximum
water absorption''. Differences in moisture content over time since saturation
can indicate different abilities to retain moisture. Leaf morphology,
especially leaf cuticle thickness, can influence water absorption
\citep{Van_Wagner-1969}. Litter particle geometry may have effects on maximum
moisture content because a larger total surface area will increase water
adsorption. Drying rate should also be influenced by leaf and litter bed traits
because surface-area-to-volume ratio and litter bulk density will influence air
flow through the litter bed. Therefore, we expect litter packing to effect fire
behavior in two ways: directly by its effect on oxygen availability for
combustion, and indirectly by its effect on drying rate. Both these pathways
are affected by leaf particle traits, especially leaf size
\citep{Scarff+Westoby-2006}. Fuel-beds composed of larger leaves or leaves with
thin cuticles should favor moisture absorption. On the other hand, large leaves
lower litter density and should favor drying, thus increasing the probability
of ignition and fire spread. The interaction of mositure absorption and drying
deserves investigation.

@@

\begin{figure}[h]
  \centering
\includegraphics[width=8cm]{figs_tables/fig1_drydown-curves.pdf}
\caption{Dry down curves for eight litter types. Lines show best-fit
  exponential decay curves fit to moisture content on a dry mass basis. Litter
  samples we saturated by immersion for 24 hours and then allowed to drain by
  gravity for 3 min before initial weighing. Colors indicate genera and there
  were significant differences in both maximum water absorption and in rate of
  drying (based linear model on log of water content). A small amount of
  horizontal jitter (0.5 hr) was added to this plot because, although
  collections were staggered, there were sometimes multiple samples collected
  at a single time point.}
 \label{fig:drydown}
\end{figure}

@@

Flame spread rate declined with fuel moisture (Fig.~\ref{fig:spread_moist}).
\emph{Quercus} had higher spread rates across all fuel moistures and had a
steeper decline in response to moisture than did the other groups (taxon by
time since wetting interaction $P = 0.022$, Table~\ref{tab:spreadrate_anova}).
Fuel moisture negatively influenced fuel consumption. Fuel consumption and the
rate of change with moisture content varied across the taxonomic groups
(Fig.~\ref{fig:consume_moist}, Table~\ref{tab:consume_anova}). Although the
overall ranking of groups was similar for fuel consumption and for flame spread
rate, there were different pattern across taxa in these two flammability
parameters. For spread rate, the decline in flammability with moisture was
shallowest for \emph{Pinus} and Cupressaceae (Fig.~\ref{fig:spread_moist}). For
fuel consumption, \emph{Pinus} and \emph{Abies} has a steeper decline in
flammability with fuel moisture than did the other groups
(Fig.~\ref{fig:consume_moist}).

%% spread rate by moisture
\begin{figure}[h]
  \centering
\includegraphics[width=8cm]{figs_tables/fig2_spread_actualMC.pdf}
\caption{Flame spread rate by moisture content (model fit $F = 61.15$, df = 9,
  148, $P < 0.0001$). Model ANOVA table in Table \ref{tab:spreadrate_anova} and
  coefficients in SI Table 4.}
  \label{fig:spread_moist}
\end{figure}

@@

%% fuel consumption by moisture:
\begin{figure}[h]
  \centering
\includegraphics[width=8cm]{figs_tables/fig3_consume_actualMC.pdf}
\caption{Percent fuel consumed by moisture content. Lines indicate best fit
  beta regression lines. Model deviance table in Table \ref{tab:consume_anova}
  and coefficients are in SI Table 5.}
  \label{fig:consume_moist}
\end{figure}

@@

As a consequence of the relationships between fuel moisture and flammability,
time since wetting had a significant effect on spread rate ($P < 0.0001$) and
the slope of this effect varied by taxonomic group (Fig. \ref{fig:spread_time},
Table~\ref{tab:spreadrate_anova}). The pines were able to ignite and burn much
sooner after wetting than did the other taxa, none of which had measurable
spread rates until burning was attempted 72 hours after wetting. Trials that
failed to even ignite smoldering combustion are omitted completely from the
model and figure. At 72 hours, the \emph{Abies}, \emph{Calocedrus}, and
\emph{Sequoiadendron} had dryer litter than did the \emph{Quercus}, yet the
spread rates were slow. \emph{Quercus} had the steepest response to time since
wetting (Fig.~\ref{fig:spread_time}, SI Table~6)).

\begin{figure}[h]
  \centering
\includegraphics[width=8cm]{figs_tables/fig4_spread_time.pdf}
\caption{Flame spread rate by time since wetting for litter of eight species.
  Lines indicate best fit linear model fits. Model ANOVA table in Table
  \ref{tab:spreadrate_anova} and coefficients in SI Table 6. Horizontal jitter
  (up to 2 hr) was added to the plot to avoid overplotting. All collections
  were at the times labeled on the x-axis. A lack of data at earlier time
  points for a taxon indicates no ignitions occurred at those time points.}
\label{fig:spread_time}
\end{figure}

Fuel consumption was highly variable across trials and intermediate values were
less common than were very low or moderately high consumption
(Fig.~\ref{fig:consume_time}). All parameters significantly contributed to the
generalized linear mixed beta regression model as indicated by the nested model
comparisons in Table~\ref{tab:consume_anova}. As was the pattern for flame
spread rate, pines had rapid flammability increases with drying. However,
unlike in the other flammability analyses, here the pines did not all behave
similarly: \emph{P. lambertiana} had lower fuel consumption than did \emph{P.
  ponderosa} and \emph{P. jeffreyi}. We noticed this during trials and in model
fitting we found that placing \emph{P. lambertiana} in its own group allowed
model convergence that was otherwise impossible. This was at odds with our
intent to use only \emph{a priori} taxonomic groups, but we could not obtain
model convergence otherwise. This result should therefore be seen as
preliminary because the test does not reflect an \emph{a priori} hypothesis,
but probably reflects the fact that \emph{P. lambertiana} has shorter and finer
needles than the other two pines.

\begin{figure}[h]
  \centering
\includegraphics[width=8cm]{figs_tables/fig5_consume_time.pdf}
\caption{Percent fuel consumed by time since wetting for litter of eight
  species. Curves shown indicate best fit beta regression lines. Model ANOVA
  table in Table \ref{tab:consume_anova} and coefficients in SI Table 7.
  Horizontal jitter (up to 2 hr) was added to the plot to avoid overplotting.
  All collections were at the times labeled on the x-axis. A lack of data at
  earlier time points for a taxon indicates no ignitions occurred at those time
  points.}
  \label{fig:consume_time}
\end{figure}

@@

\begin{figure}[h]
  \centering
\includegraphics[width=16cm]{figs_tables/fig6_mixture_obs_vs_pred_mc.pdf}
\caption{Observed vs predicted fuel moisture content for litter mixtures.
  Predictions are based on the mean moisture content at each time since wetting
  for the three constituent species in each mixture. Codes for species in
  mixtures: Ab = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi =
  \emph{Pinus jeffreyi} and Qu = \emph{Quercus kellogii}.}
  \label{fig:mixture_obs_pred_mc}
\end{figure}

@@

\begin{figure}[h]
  \centering
\includegraphics[width=16cm]{figs_tables/fig7_mixture_obs_vs_pred_spread.pdf}
\caption{Observed vs. predicted flame spread rate for litter mixtures. Filled
  circles show observed spread rate values. Open diamonds indicate predicted
  spread rates based on the average spread rate at each time since wetting for
  the three constituent species in each mixture. Codes for species in mixtures:
  Ab = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi =
  \emph{Pinus jeffreyi} and Qu = \emph{Quercus kellogii}.}
  \label{fig:mixture_obs_pred_spread}
\end{figure}

\begin{figure}[h]
  \centering
\includegraphics[width=16cm]{figs_tables/fig8_mixture_obs_vs_pred_consume.pdf}
\caption{Observed vs predicted fuel consumption for litter mixtures. Filled
  circles show observed fuel consumption percentages. Open diamonds indicate
  predicted fuel consumption based on the average fuel consumption at each
  particular time since wetting for the three constituent species in each
  mixture. Codes for species in mixtures: Ab = \emph{Abies concolor}, Ca =
  \emph{Calocedrus decurrens}, Pi = \emph{Pinus jeffreyi} and Qu =
  \emph{Quercus kellogii}.}
  \label{fig:mixture_obs_pred_consume}
\end{figure}

@@

The taxa in this study displayed different patterns of both moisture absorption
and desorption, associated with different litter traits. In this sample of
species, those that have a greater absorptive capacity are also among the most
flammable in terms of spread rate when dry. Traits such as large leaves and
high SLA may affect both absorptive capacity and flammability with a
consequence that flammability rankings across species can change with time
since wetting. One shortcoming of past work in this system was that
flammability trials tested only dry fuels \citep{Magalhaes+Schwilk-2012}. We
therefore expected that rankings of species flammability might change under
higher moisture levels. For example, one presumption of ours prior to this
experiment was that the high flammability previously reported for \emph{Q.
  kelloggii} \citep{Magalhaes+Schwilk-2012} would disappear under higher fuel
moisture contents. This prediction was partly supported: \emph{Q. kelloggii}
fails to ignite until after 72 hours of drying in this study (Figs.
\ref{fig:spread_time} and \ref{fig:consume_time}). However, high moisture
content in this oak was unable to suppress ignition of mixtures to which it
contributed (Figs. \ref{fig:mixture_obs_pred_spread} and
\ref{fig:mixture_obs_pred_consume}).

@@

Our flammability trials demonstrate that both flame spread rate and fuel
consumption increased with time since wetting but that the rate of increase
varied across these taxa. Only the pines are able to ignite and support flame
spread immediately after wetting (Fig. \ref{fig:spread_time}). In contrast,
\emph{Quercus} exhibited a threshold response to drying with no ignition prior
to 72 hours of drying but once ignition is possible, it had faster spread rates
than the other taxa. Our two Cupressaceae, which have dense litter but have low
maximum water absorption, had low flammability across moisture levels. This
result is consistent with the low flammability found in an Australian
Cupressaceae that have small litter particles and dense packing
\citep{Scarff+Westoby-2006, Bowman+Haverkamp+etal-2018}. However, the densely
packed litter is not a consistent trait across the family because
\citet{Cornwell+Elvira+etal-2015} found relatively high litter flammability and
low litter density for some Asian Cupressaceae.

>>>>

\documentclass[letterpaper,12pt]{article}
\usepackage{jecol}

@@

\nwords{8200}
\ntables{4}
\nfig{5}
\nref{55}

@@

Fuel moisture content through time is controlled by two processes, wetting
(including absorption and adsorption) and drying rate (desorbtion)
\citep{Anderson-1990, Kreye+Kobziar+etal-2013}. Absorptive capacity represents
the potential maximum fuel moisture and is evaluated by the moisture content of
litter when saturated; hereafter we refer to this measurement as ``maximum
water absorption''. Differences in moisture content over time since saturation
can indicate different abilities to retain moisture. Leaf morphology,
especially leaf cuticle thickness, can influence water absorption
\citep{Van_Wagner-1969}. Litter particle geometry may have effects on maximum
moisture content because a larger total surface area will increase water
adsorption. Drying rate should also be influenced by leaf and litter bed traits
because surface-area-to-volume ratio and litter bulk density will influence air
flow through the litter bed. Therefore, we expect litter packing to effect fire
behavior in two ways: directly by its effect on oxygen availability for
combustion, and indirectly by its effect on drying rate. Both these pathways
are affected by leaf particle traits, especially leaf size
\citep{Scarff+Westoby-2006}. Fuel-beds composed of larger leaves or leaves with
thin cuticles should favor moisture absorption. On the other hand, large leaves
lower litter density and should favor drying, thus increasing the probability
of ignition and fire spread. The interaction of moisture absorption and drying
deserves investigation.

@@

\begin{figure}[h]
  \centering
\includegraphics[width=5cm]{figs_tables/fig1_dry_effects_by_taxon.pdf}
\caption{Effects of litter drying on fuel moisture and flammability. A) Dry
  down curves for eight species. Lines show best-fit exponential decay curves
  fit to moisture content on a dry mass basis. B) Flame spread rate by time
  since wetting. Lines indicate best fit linear model fits by taxonomic group.
  Model ANOVA table in Table \ref{tab:spreadrate_anova} and coefficients in SI
  Table 6. C) Percent fuel consumed by time since wetting. Curves shown
  indicate best fit beta regression lines by taxonomic group. Model ANOVA table
  in Table \ref{tab:consume_anova} and coefficients in SI Table 7. For all
  plots, horizontal jitter (0.5 to 2 hr) was added to avoid overplotting. A
  lack of data at earlier time points for a taxon indicates no ignitions
  occurred at those time points. }
 \label{fig:drydown}
\end{figure}

@@

Flame spread rate declined with fuel moisture (Fig.~\ref{fig:flam_moist}).
\emph{Quercus} had higher spread rates across all fuel moistures and had a
steeper decline in response to moisture than did the other groups (taxon by
time since wetting interaction $P = 0.022$, Table~\ref{tab:spreadrate_anova}).
Fuel moisture negatively influenced fuel consumption. Fuel consumption and the
rate of change with moisture content varied across the taxonomic groups
(Fig.~\ref{fig:flam_moist}, Table~\ref{tab:consume_anova}). Although the
overall ranking of groups was similar for fuel consumption and for flame spread
rate, there were different pattern across taxa in these two flammability
parameters. For spread rate, the decline in flammability with moisture was
shallowest for \emph{Pinus} and Cupressaceae (Fig.~\ref{fig:flam_moist}). For
fuel consumption, \emph{Pinus} and \emph{Abies} has a steeper decline in
flammability with fuel moisture than did the other groups
(Fig.~\ref{fig:flam_moist}).

%% spread rate by moisture
\begin{figure}[h]
  \centering
\includegraphics[width=8cm]{figs_tables/fig2_moisture_flam.pdf}
\caption{Fuel moisture effects on flammability. A) Flame spread rate by moisture content (model fit $F = 61.15$, df = 9,
  148, $P < 0.0001$). Model ANOVA table in Table \ref{tab:spreadrate_anova} and
  coefficients in SI Table 4. B) Percent fuel consumed by moisture content. Lines indicate best fit
  beta regression lines. Model deviance table in Table \ref{tab:consume_anova}
  and coefficients are in SI Table 5.}
  \label{fig:flam_moist}
\end{figure}

@@

@@

As a consequence of the relationships between fuel moisture and flammability,
time since wetting had a significant effect on spread rate ($P < 0.0001$) and
the slope of this effect varied by taxonomic group (Fig. \ref{fig:drydown},
Table~\ref{tab:spreadrate_anova}). The pines were able to ignite and burn much
sooner after wetting than did the other taxa, none of which had measurable
spread rates until burning was attempted 72 hours after wetting. Trials that
failed to even ignite smoldering combustion are omitted completely from the
model and figure. At 72 hours, the \emph{Abies}, \emph{Calocedrus}, and
\emph{Sequoiadendron} had drier litter than did the \emph{Quercus}, yet the
spread rates were slow. \emph{Quercus} had the steepest response to time since
wetting (Fig.~\ref{fig:drydown}, SI Table~6)).

Fuel consumption was highly variable across trials and intermediate values were
less common than were very low or moderately high consumption
(Fig.~\ref{fig:drydown}). All parameters significantly contributed to the
generalized linear mixed beta regression model as indicated by the nested model
comparisons in Table~\ref{tab:consume_anova}. As was the pattern for flame
spread rate, pines had rapid flammability increases with drying. However,
unlike in the other flammability analyses, here the pines did not all behave
similarly: \emph{P. lambertiana} had lower fuel consumption than did \emph{P.
  ponderosa} and \emph{P. jeffreyi}. We noticed this during trials and in model
fitting we found that placing \emph{P. lambertiana} in its own group allowed
model convergence that was otherwise impossible. This was at odds with our
intent to use only \emph{a priori} taxonomic groups, but we could not obtain
model convergence otherwise. This result should therefore be seen as
preliminary because the test does not reflect an \emph{a priori} hypothesis,
but probably reflects the fact that \emph{P. lambertiana} has shorter and finer
needles than the other two pines.

@@

\begin{figure}[h]
  \centering
\includegraphics[width=16cm]{figs_tables/fig3_mixture_obs_vs_pred_mc.pdf}
\caption{Observed vs predicted fuel moisture content for litter mixtures.
  Predictions are based on the mean moisture content at each time since wetting
  for the three constituent species in each mixture. Codes for species in
  mixtures: Ab = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi =
  \emph{Pinus jeffreyi} and Qu = \emph{Quercus kellogii}.}
  \label{fig:mixture_obs_pred_mc}
\end{figure}

@@

\begin{figure}[h]
  \centering
\includegraphics[width=16cm]{figs_tables/fig4_mixture_obs_vs_pred_spread.pdf}
\caption{Observed vs. predicted flame spread rate for litter mixtures. Filled
  circles show observed spread rate values. Open diamonds indicate predicted
  spread rates based on the average spread rate at each time since wetting for
  the three constituent species in each mixture. Codes for species in mixtures:
  Ab = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi =
  \emph{Pinus jeffreyi} and Qu = \emph{Quercus kellogii}.}
  \label{fig:mixture_obs_pred_spread}
\end{figure}

\begin{figure}[h]
  \centering
\includegraphics[width=16cm]{figs_tables/fig5_mixture_obs_vs_pred_consume.pdf}
\caption{Observed vs predicted fuel consumption for litter mixtures. Filled
  circles show observed fuel consumption percentages. Open diamonds indicate
  predicted fuel consumption based on the average fuel consumption at each
  particular time since wetting for the three constituent species in each
  mixture. Codes for species in mixtures: Ab = \emph{Abies concolor}, Ca =
  \emph{Calocedrus decurrens}, Pi = \emph{Pinus jeffreyi} and Qu =
  \emph{Quercus kellogii}.}
  \label{fig:mixture_obs_pred_consume}
\end{figure}

@@

The taxa in this study displayed different patterns of both moisture absorption
and desorption, associated with different litter traits. In this sample of
species, those that have a greater absorptive capacity are also among the most
flammable in terms of spread rate when dry. Traits such as large leaves and
high SLA may affect both absorptive capacity and flammability with a
consequence that flammability rankings across species can change with time
since wetting. One shortcoming of past work in this system was that
flammability trials tested only dry fuels \citep{Magalhaes+Schwilk-2012}. We
therefore expected that rankings of species flammability might change under
higher moisture levels. For example, one presumption of ours prior to this
experiment was that the high flammability previously reported for \emph{Q.
  kelloggii} \citep{Magalhaes+Schwilk-2012} would disappear under higher fuel
moisture contents. This prediction was partly supported: \emph{Q. kelloggii}
fails to ignite until after 72 hours of drying in this study (Figs.
\ref{fig:drydown} and \ref{fig:drydown}). However, high moisture
content in this oak was unable to suppress ignition of mixtures to which it
contributed (Figs. \ref{fig:mixture_obs_pred_spread} and
\ref{fig:mixture_obs_pred_consume}).

@@

Our flammability trials demonstrate that both flame spread rate and fuel
consumption increased with time since wetting but that the rate of increase
varied across these taxa. Only the pines are able to ignite and support flame
spread immediately after wetting (Fig. \ref{fig:drydown}). In contrast,
\emph{Quercus} exhibited a threshold response to drying with no ignition prior
to 72 hours of drying but once ignition is possible, it had faster spread rates
than the other taxa. Our two Cupressaceae, which have dense litter but have low
maximum water absorption, had low flammability across moisture levels. This
result is consistent with the low flammability found in an Australian
Cupressaceae that have small litter particles and dense packing
\citep{Scarff+Westoby-2006, Bowman+Haverkamp+etal-2018}. However, the densely
packed litter is not a consistent trait across the family because
\citet{Cornwell+Elvira+etal-2015} found relatively high litter flammability and
low litter density for some Asian Cupressaceae.
